Staged Abstract Interpreters

### 2019-09-04 paper/20_prelim.tex

As the first step in our roadmap, we instantiate the interpreter concretely in
this section. The result of such instantiation is a standard definitional
interpreter with environments and stores, it also can be obtained by
refunctionalization of CESK machines \cite{Felleisen:1987:CAH:41625.41654,
DBLP:conf/ppdp/AgerBDM03}. We first present the concrete components, i.e., the
value domain, then show the monad stack for concrete interpretation, finally
sketch the how the primitive operations are implemented.

\paragraph{Concrete Components}
The two types we need to concretize are addresses @Addr@ and values @Value@. The
type @Env@ and @Store@ is derived automatically. To assure the freshness of
address allocation, we use @Int@ and always return successor of the size of
current store. A value can be either a tagged number \texttt{IntV}, or a closure
\texttt{CloV} that contains a lambda term and an environment. The final grounded
result of the interpreter is a pair of values and stores, where the @Value@ and
@Store@ are next-stage objects, as annotated by @R@. We also define a standard
fixed-point combinator to close the open-recursive function @ev@.
\begin{lstlisting}
  trait ConcreteComponents extends Semantics {
    type Addr = Int
    sealed trait Value
    case class IntV(i: Int) extends Value
    case class CloV(λ: Lam, e: Env) extends Value
    type Result = (R[Value], R[Store])
    def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => ev(fix(ev))(e)
  }
\end{lstlisting}

\paragraph{Unstaged Monads}
For the purpose of concrete interpretation, the monad needs to model read
effect and state effect, which correspond to environment and store,
respectively. Following the monad transformer approach
\cite{DBLP:conf/popl/LiangHJ95}, we use a @ReaderT@ and @StateT@ monad
transformer to compose the monad stack. In other words, the type @AnsM@ is
instantiated by layering @ReaderT@ and @StateT@ transformers\footnote{The
question mark syntax is a kind projector \cite{kindprojector}, thus
\texttt{StateT[IdM,Store,?]} is equivalent to \newline \texttt{(\{type
M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is parameterized by type
@Env@, @StateT@ is parameterized by type @Store@, and the inner-most monad @IdM@
is simply an identity monad.
\begin{lstlisting}
  trait ConcreteSemantics extends ConcreteComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[IdM, Store, ?], Env, T] // the monad stack
    ... }
\end{lstlisting}

Here we sketch the basic idea of @ReaderT@ and @StateT@. Readers may
refer to \cite{DBLP:conf/popl/LiangHJ95, Chiusano:2014:FPS:2688794} for more detail.
A @ReaderT@ monad transformer encapsulates computation @R => M[A]@, where
@R@ is the reader type, and @M[_]@ is the inner monad type.
Given a @R@, @ReaderT@ monad produces a transformed value of type @M[A]@.
Similarly, a @StateT@ monad encapsulates computation @S => M[(A, S)]@, where
@S@ is the state type, and @M[_]@ is the inner monad type.
Given a @S@, @StateT@ monad produces a transformed value of type @M[(A, S)]@,
where the new state (type @S@) is accompanied with the value (type @A@).
Note that for the moment, the binding-time type @R@ is an identity type, thus
these monads operate on unstaged data. We can also see this from the
signature of @flatMap@: the argument function @f@ takes an unstaged value of type @A@ and
produces a monadic value. In the following code, we elide operations other than @flatMap@.
\begin{lstlisting}
  case class ReaderT[M[_]: Monad, R, A](run: R => M[A]) {
    def flatMap[B](f: A => ReaderT[M, R, B]): ReaderT[M, R, B] =
      ReaderT(r => Monad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class StateT[M[_]: Monad, S, A](run: S => M[(A, S)]) {
    def flatMap[B](f: A => StateT[M, S, B]): StateT[M, S, B] =
      StateT(s => Monad[M].flatMap(run(s)) { case (a, s1) => f(a).run(s1) }); ... }
\end{lstlisting}

After defining the monad stack, the operations that manipulate the environment
and store can be defined by constructing the proper monad and lifting it to the
top-level of our monad stack. To modify the store, for example, we can construct a
@StateT@ value that transforms the current store $\sigma$ to @σ + αv@, which
results in @StateT[IdM, Store, Unit]@, and then lift this @StateT@ value to the
top-level @ReaderT@ type, i.e., @AnsM[Unit]@.
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ + αv))
\end{lstlisting}

\paragraph{Primitive Operations}
Other primitive operations over the value domain can be implemented
straightforwardly. We elide most of them but describe a bit on how we handle
functions and applications. Because lambda terms are data but also part of the
control flow, although the way we treat them is simple now, but will be very
different when staging and abstract interpretation is involved. The method
@close@ takes a lambda term and an environment, and produces the
defunctionalized representation of closures.
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = CloV(λ, ρ)
\end{lstlisting}

To apply a function, @ap_clo@ takes a function value @rator@ and an argument
@rand@, extracts the lambda term and environment enclosed in @rator@, allocates
the addresses and updates the environment and store accordingly. Finally, it
evaluates the body expression of the lambda term under the new environment and
store.
\begin{lstlisting}
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = rator match {
    case CloV(Lam(x, e), ρ: Env) => for {
      α <- alloc(x)
      _ <- set_store(α → rand)
      rt <- local_env(ev(e))(ρ + (x → α))
    } yield rt
  }
\end{lstlisting}

>>>>

As the first step in our roadmap, we instantiate the interpreter concretely in
this section. The result of such instantiation is a standard definitional
interpreter with environments and stores. It can also be obtained by the
refunctionalization of CESK machines \cite{Felleisen:1987:CAH:41625.41654,
DBLP:conf/ppdp/AgerBDM03}. We first present the concrete components, i.e., the
value domains, then show the monad stack for concrete interpretation, finally
sketch how the primitive operations are implemented.

\paragraph{Concrete Components}
The two types we need to concretize are addresses @Addr@ and values @Value@. The
type @Env@ and @Store@ is derived automatically. To assure the freshness of
address allocations, we use type @Int@ and always return a number greater
than the size of current store.
A value can be either a tagged number \texttt{IntV}, or a closure
\texttt{CloV} that contains a lambda term and an environment. The final grounded
result of the interpreter is a pair of values and stores, where the @Value@ and
@Store@ can be next-stage objects, as annotated by the binding-time @R@. We
also define a standard fixed-point combinator to close the open-recursive
function @ev@.
\begin{lstlisting}
  trait ConcreteComponents extends Semantics {
    type Addr = Int
    sealed trait Value
    case class IntV(i: Int) extends Value
    case class CloV(λ: Lam, e: Env) extends Value
    type Result = (R[Value], R[Store])
    def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => ev(fix(ev))(e)
  }
\end{lstlisting}

\paragraph{Unstaged Monads}
For the purpose of concrete interpretation, the monad needs to model read
effect and state effect, which correspond to environment and store,
respectively. Following the monad transformer approach
\cite{DBLP:conf/popl/LiangHJ95}, a @ReaderT@ and @StateT@ monad
transformer are used to compose the monad stack. In other words, the type @AnsM@ is
instantiated by layering @ReaderT@ and @StateT@ transformers\footnote{The
question mark syntax is a kind projector \cite{kindprojector}, thus
\texttt{StateT[IdM,Store,?]} is equivalent to \newline \texttt{(\{type
M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is parameterized by type
@Env@, and @StateT@ is parameterized by type @Store@, and the inner-most monad @IdM@
is simply an identity monad.
\begin{lstlisting}
  trait ConcreteSemantics extends ConcreteComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[IdM, Store, ?], Env, T] // the monad stack
    ... }
\end{lstlisting}

Here we sketch the basic idea of @ReaderT@ and @StateT@. Readers may
refer to \cite{DBLP:conf/popl/LiangHJ95, Chiusano:2014:FPS:2688794} for more detail.
A @ReaderT@ monad transformer encapsulates computation @R => M[A]@, where
@R@ is the reader type, and @M[_]@ is the inner monadic type.
Given a value of @R@, a @ReaderT@ monad produces a transformed value of type @M[A]@.
Similarly, a @StateT@ monad encapsulates computation @S => M[(A, S)]@, where
@S@ is the state type, and @M[_]@ is the inner monad type.
Given a value of @S@, a @StateT@ monad produces a transformed value of type @M[(A, S)]@,
where the new state (of type @S@) is accompanied with the value (of type @A@).
Note that for the moment, the binding-time type @R@ is an identity type, thus
these monads operate on unstaged data. We can also see this from the
signature of @flatMap@: the argument function @f@ takes an unstaged value of type @A@ and
produces a monadic value. In the following code, we elide operations other than @flatMap@.
\begin{lstlisting}
  case class ReaderT[M[_]: Monad, R, A](run: R => M[A]) {
    def flatMap[B](f: A => ReaderT[M, R, B]): ReaderT[M, R, B] =
      ReaderT(r => Monad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class StateT[M[_]: Monad, S, A](run: S => M[(A, S)]) {
    def flatMap[B](f: A => StateT[M, S, B]): StateT[M, S, B] =
      StateT(s => Monad[M].flatMap(run(s)) { case (a, s1) => f(a).run(s1) }); ... }
\end{lstlisting}

After defining the monad stack, the operations manipulating environments
and stores can be defined by constructing the proper monadic value and lifting it to the
top-level of our monad stack. To modify the store, for example, we construct a
@StateT@ value that transforms the current store $\sigma$ to @σ + αv@, which
results in a @StateT[IdM, Store, Unit]@, and then lift this @StateT@ value to the
top-level @ReaderT@ type, i.e., @AnsM[Unit]@.
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ + αv))
\end{lstlisting}

\paragraph{Primitive Operations}
Other primitive operations over the value domain can be implemented
straightforwardly. We elide most of them but describe a bit on how we handle
functions and applications. This is because lambda terms are data but are also
part of the control flow. Although the way we treat them is simple now, but it
will be very different when staging and abstract interpretation is involved.
At the moment, the method @close@ takes a lambda term and an environment, and
produces the defunctionalized representation of closures.
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = CloV(λ, ρ)
\end{lstlisting}

To apply a function, @ap_clo@ takes a function value @rator@ and an argument
@rand@. It extracts the lambda term and environment enclosed in @rator@, and
allocates the addresses and updates the environment and store accordingly.
Finally, body expression of the lambda term is evaluated under the new
environment and store.
\begin{lstlisting}
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = rator match {
    case CloV(Lam(x, e), ρ: Env) => for {
      α <- alloc(x)
      _ <- set_store(α → rand)
      rt <- local_env(ev(e))(ρ + (x → α))
    } yield rt
  }
\end{lstlisting}
